---
title: "Building on ggplot2 for exploration of missing values"
author: "Nicholas Tierney"
output:
  pdf_document:
    keep_tex: yes
  html_document:
    keep_md: yes
bibliography: jsm2017.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dev = "png",
                      dpi = 200)
```

```{r library-calls, echo = FALSE, include = FALSE}

# install ggplot2 version 2.1.0
# # devtools::install_version("ggplot2","2.1.0")
library(tidyverse)
library(ggmissing)
library(visdat)
```

```{r narnia-functions, include = FALSE}


# give NAs a more meaningful label
is_na <- function(x) {
  factor(is.na(x), levels = c(FALSE, TRUE), labels = c("!NA", "NA"))
}

# append some shadow cols

# return a tibble that is a shadow matrix form.

as_shadow <- function(data){

  data_shadow <- map_df(data, is_na)

  names(data_shadow) <- paste0(names(data),"_NA")

  data_shadow

  }

# as_shadow(airquality)

bind_shadow <- function(data){

  # data_shadow <- map_df(data, is_na)
  # names(data_shadow) <- paste0(names(data),"_NA")
  data_shadow <- as_shadow(data)

  bound_shadow <- as_tibble(dplyr::bind_cols(data, data_shadow))

  bound_shadow

}

gather_shadow <- function(data){
  
  as_shadow(data) %>%
  mutate(rows = 1:nrow(.)) %>%
  gather(key = "var",
         value = "miss",
         -rows) 
}

prop_na <- function(x){
  
  prop_na_df <- data_frame(
    df = percent_missing_df(x),
    var = percent_missing_var(x),
    case = percent_missing_case(x)
  )
  
  prop_na_df
  
}

# prop_na(airquality)
# tao1 <- tao %>% filter(year == 1997)
# tao2 <- tao %>% filter(year == 1993)
# # 
# ggmissing::summary_missing_var(tao)
# ggmissing::summary_missing_var(tao1)
# ggmissing::summary_missing_var(tao2)
# # 

# percent_missing_df()
# miss_prop_df(tao)
# prop_na_df(tao)
# mean(is.na(tao))


```

# Introduction

Missing values are ubiquitous in data and need to be explored and handled in the initial stages of analysis ggplot2, an implementation of the grammar of graphics is an incredibly popular way to produce data visualisations does not currently support missing data [@wickham2009ggplot2]. Principles of tidy data [@wickham2014] states that each row is an observation and each column is a variable, which makes it easy and consistent to perform data manipulation and wrangling. However, there are currently no guidelines for representing additional missing data structures in a tidy format. This paper describes approaches for exploring missing data structure with minimal deviation from the common workflows of ggplot and tidy data structures.

# Missing Data Mechanisms

Canonical sources of missing data are questionnaires. Data obtained from questionnaires are often subject to both unknown and known missingness structure. Unknown missing data structure may arise from respondents accidentally failing to answer questions or inadvertently providing inappropriate answers. Known missing data structure data may arise due to the structure of the questionnaire. For example, the first question on a survey might be: 'If YES, skip to question 4', resulting in questions 2 and 3 missing. If the structure of the questionnaire is known, this type of missingness can be evaluated easily. However, if this information is not available, the mechanism responsible for producing missing data must be inferred from the data. Longitudinal studies are also sources of missing data, where participants may not return for future testing sessions. In these cases it is difficult, sometimes impossible, to ascertain the reason for the dropouts, and hence, whether the missingness structure is known or unknown.

There are a two approaches to analysis of data with missing values, deletion and imputation. Deletion methods drop variables or cases, depending on the amount of missing data, and imputation methods replace the missing values with some other value estimated from the data. It is now widely regarded as best practice to impute these values, however in order for estimates to be unbiased, it is essential to understand the missingness structure and mechanisms [@Little1988; @Rubin1976; @simon-simonoff; @Schafer2002]. 

## Existing packages for handling missing data

Software focussing on missing data typically focus on imputation or visualisation. Packages such as mice, mi, norm, and Amelia provide functions to facilitate imputation, and use a wide range of methods, from mean or median imputation, to regression or machine learning, to Bayesian methodologies, as well as providing diagnostics on the imputations themselves [@mice; @mi; @norm; @Amelia].

Missing data visualisation packages include the R package VIM, and the stand alone softwares MANET, ggobi, MissingDataGUI, and to a more limited extent, ggplot2 [@cheng2015; @Unwin1996; @swayne2003ggobi; @vim; @wickham2009ggplot2].  MANET (Missings Are Now Equally Treated), provides univariate visualisations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. ggobi extends the univariate linked brushing of MANET to multivariate, using parallel co-ordinate plots. ggobi also provided incoporated missingness into scatterplots, displaying missing values from one variable as 10% below the minimum value on the other axis. MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually. Using a GUI to explore missing data makes it easier to glean valuable insights into important structures, but may then make it hard to incorporate these unscripted insights into reproducible analyses, and may also distract and break the workflow from statistical analysis.

VIM (Visualising and Imputing Missing Data) is an R package that provides methods for both imputation and visualisation of missing data. In particular it provides visualisations that identify observed, imputed, and missing values. VIM also identifies imputed cases by adding a suffix to a variable, so Var1 would have a sibling indicator column, Var1_imp, where each case is TRUE or FALSE to indicate imputation. Although VIM provides R functions to visualise and impute missing data, it's syntax for data manipulation and visualisation is difficult to extend, and does not follow tidy data principles. ggplot2 currently only provides visualisation of missing values for categories treating categories as NA values. For all other plots, ggplot2 prints a warning message of the number of missing values ommited.

There are many ways to explore missing data structure and imputation, however there is no unified methodology to explore, or visualise missing data. We now discuss ways to represent missing data that fit in with the grammar of graphics and tidy data.

# Data structures for missing data

Representing missing data structure is achieed using the shadow matrix, introduced in Swayne and Buja @Swayne1998 [-@Swayne1998]. The shadow matrix is the same dimension as the data, and consists of binary indicators of missingness of data values, where missing is represented as "NA", and not missing is represented as "!NA". Although these may be represented as 1 and 0, respectively. This representation can be seen in figure 1 below, adding the sufic "_NA" to the variables. This structure can also be extended to allow for additional factor levels to be created. For example 0 indicates data presence, 1 indicates missing values, 2 indicates imputed value, and 3 might indicate a particular type or class of missingness, where reasons for missingness might be known or inferred. The data matrix can also be augmented to include the shadow matrix, which facilitates visualisation of univariate and bivariate missing data visualisations. Another format is to display it in long form, which facilitates heatmap style visualisations. This approach can be very helpful for giving an overview of which variables contain the most missingness. Methods can also be applied to rearrange rows and columns to find clusters, and identify other interesting features of the data that may have previously been hidden or unclear. 

\begin{figure}[h]
\centering
\includegraphics[width=270pt]{diagram.png}
\end{figure}

__Figure 1. Illustration of data structures for facilitating visualisation of missings and not missings.__

# Visualising missing data

**Heatmap**

A missing data heatmap is shown below using the `vis_miss` command from the `visdat` package. This displays the the airquality dataset included in base R, which contains Daily air quality measurements in New York, May to September 1973.


```{r fig.height = 4, fig.width = 10, echo = FALSE}

library(visdat)
p1 <- vis_miss(airquality)
p2 <- vis_miss(airquality, cluster = TRUE)

gridExtra::grid.arrange(p1,p2,ncol = 2)

```

__Figure 2. Heatmaps of missing data (left) raw (right) ordered by clustering on rows and columns.__

Similar approaches have been used in other missing data packages such as VIM, mi, Amelia, and MissingDataGUI. However this plot is created in the ggplot framework, giving users greater control over the plot appearance. The user can also apply clustering of the rows and columns using the `cluster = TRUE` argument (shown on the right).

**Univariate plots split by missingness**

An advantage of the augmented shadow format, where the data and shadow are side by side, is that it allows for examining univariate distributions according to the presence or absence of another variable. The plot below shows the values of temperature when ozone is present and missing, on the left is a faceted histogram, and on the right is an overlayed density.

```
ggplot(data = bind_shadow(airquality),
       aes(x = Temp)) + 
  geom_histogram() + 
  facet_wrap(~Ozone_NA,
             ncol = 1)

ggplot(data = bind_shadow(airquality),
       aes(x = Temp,
           colour = Ozone_NA)) + 
  geom_density()
```



```{r bind-shadow-density, fig.height = 4, fig.width = 10, error = FALSE, message = FALSE, echo = FALSE, }


p1 <- ggplot(data = bind_shadow(airquality),
       aes(x = Temp)) + 
  geom_histogram(na.rm = TRUE) + 
  facet_wrap(~Ozone_NA,
             ncol = 1)

p2 <- ggplot(data = bind_shadow(airquality),
       aes(x = Temp,
           colour = Ozone_NA)) + 
  geom_density(na.rm = TRUE) 

gridExtra::grid.arrange(p1,p2,ncol = 2)

```

__Figure 3., Two representations of temperature showing missingness on ozone (left) facetted, (right) coloured. Main difference in the distribution is that there is a cluster of low temperature observations with missing ozone values.__

Using this data structure allows for the user to directly refer to the variable for which they want to explore the effect of missingness using the suffix _NA after the variable. In the case above, the user is looking at a histogram of temperature, but is then able to look at how many temperature values are affected by missingness of ozone. In cases where there is no missing data in the variable that they want to "split" the missingness by, the plot simple returns a single facetted plot.

Another method of visualisation can be explored using `geom_missing_point()` from the `ggmissing` package:

```{r ggeom_missing, fig.height = 4, fig.width = 8}

p1 <- ggplot(data = airquality,
       aes(x = Ozone,
           y = Solar.R)) + 
  geom_missing_point() + 
  theme(aspect.ratio = 1)

p2 <- ggplot(data = airquality,
       aes(x = Temp,
           y = Ozone)) + 
  geom_missing_point() + 
  theme(aspect.ratio = 1)

gridExtra::grid.arrange(p1,p2,ncol=2)
```

__Figure 4., Scatterplots with missings displayed at 10% below (left) ozone versus and solar radiation (right) ozone versus temperature.__

This replaces missing values to be 10% below the minimum value, a technique borrowed from ggobi. The missing values are also different colours to make missingness preattentive [@treisman1985]. In this plot we see that there is a mostly uniform spread of missing values for Solar.R and Ozone. As `geom_missing_point` is a defined geometry for ggplot2, it allows users to have full customisation as they normally would with ggplot.

# Numerical Summaries for missing data

Numerical summaries of missing data are also made easy with some helper functions from the `ggmissing` package, which provides tidy functions that return either single numbers or dataframes. The `percent_missing_*` functions help find the proportion of missing values in the data overall, in cases, or in variables. 

```{r numerical-percent-missing}

# Proportion elements in dataset that contains missing values
percent_missing_df(airquality)
# Proportion of variables that contain any missing values
percent_missing_var(airquality)
 # Proportion of cases that contain any missing values
percent_missing_case(airquality)

```

We can also look at the number and percent of missings in each case and variable with `summary_missing_case`, and `summary_missing_var`.

```{r}

summary_missing_case(airquality) %>% slice(1:5)
summary_missing_var(airquality)

```

Tabulations of the number of missings in each case or variable can be calculated with `table_missing_case` and `table_missing_var`. 

```{r}

table_missing_case(airquality)
table_missing_var(airquality)

```

<!-- it would be really cool if we could implement dplyr `group_by` syntax for the data, to produce summaries of missingness for 1993 and 1997 respectively. -->

# Discussion

In this paper we discussed missing data mechanisms, existing packages for imputation and visualisation of missing data, and the limitations of current missing data exploration and visualisaion softwares. We then discussed data structures for missing data, and showed how these can be used following tidy data principles, and how to effectively present visualisations and numerical summaries using the R packages ggmissing and visdat, available for download on github: https://github.com/njtierney/ggmissing, and https://github.com/njtierney/visdat.

It is worthwhile to note the trade off between storage and computation of the augmented shadow matrix. When storage of data is an issue, it may not be practial to bind the shadow matrix to the regular data. Instead, it may be more effective to perform the computation for the column of interest when necessary. However, the shadow matrix can also allow for more complex types of missingness to be expressed, and so there are additional benefits to storing data in this way. For example, missing, `NA`,and not missing, `!NA`, could be extended to describe different mechanisms for missingness, e.g., `NA_mechanism_A`, and `NA_mechanism_B`, or even imputed values `value_imputed`. These could then be combined with the same sorts of plots and numerical summaries to provide diagnostics.

Future research should focus on developing techniques for identifying missingness mechanisms and methods for encoding mechanisms into the shadow matrix. Further work could also be done on developing methods to store single and multiple imputations into the shadow matrix, and methods to visualise these imputations using ggplot geoms, and assess them with numerical summaries.

# References

