---
title: 'naniar: Expanding ggplot2 for better visual exploration missingsness'
author: "Nicholas Tierney, Di Cook"
date: "14/12/2016"
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dev = "png",
                      dpi = 200)
```

```{r library-calls, echo = FALSE, include = FALSE}

# install ggplot2 version 2.1.0
# # devtools::install_version("ggplot2","2.1.0")
library(tidyverse)
library(ggmissing)
library(visdat)
```

```{r narnia-functions, include = FALSE}


# give NAs a more meaningful label
is_na <- function(x) {
  factor(is.na(x), levels = c(FALSE, TRUE), labels = c("!NA", "NA"))
}

# append some shadow cols

# return a tibble that is a shadow matrix form.

as_shadow <- function(data){

  data_shadow <- map_df(data, is_na)

  names(data_shadow) <- paste0(names(data),"_NA")

  data_shadow

  }

# as_shadow(airquality)

bind_shadow <- function(data){

  # data_shadow <- map_df(data, is_na)
  # names(data_shadow) <- paste0(names(data),"_NA")
  data_shadow <- as_shadow(data)

  bound_shadow <- as_tibble(dplyr::bind_cols(data, data_shadow))

  bound_shadow

}

gather_shadow <- function(data){
  
  as_shadow(data) %>%
  mutate(rows = 1:nrow(.)) %>%
  gather(key = "var",
         value = "miss",
         -rows) 
}

prop_na <- function(x){
  
  prop_na_df <- data_frame(
    df = percent_missing_df(x),
    var = percent_missing_var(x),
    case = percent_missing_case(x)
  )
  
  prop_na_df
  
}

# prop_na(airquality)
# tao1 <- tao %>% filter(year == 1997)
# tao2 <- tao %>% filter(year == 1993)
# # 
# ggmissing::summary_missing_var(tao)
# ggmissing::summary_missing_var(tao1)
# ggmissing::summary_missing_var(tao2)
# # 

# percent_missing_df()
# miss_prop_df(tao)
# prop_na_df(tao)
# mean(is.na(tao))


```

# Introduction

Missing data is ubiquitous in data analysis, and are often the source of much energy, frustration, and confusion. Since 2014 there has been substantial development in the area of "tidy data" [ @wickham], which states the (surprisingly simple!) rule that each row is an observation and each column is a variable, which makes it easy to reason with data. This paper describes approaches for summarising missing data in numerical and graphical forms whilst maintaining a tidy format.

# Types of missing data

Canonical sources of missing data are questionnaires. Data obtained from questionnaires are often subject to both unknown and known missingness structure. For example, unknown missing data structure can arise from respondents accidentally failing to answer questions or inadvertently providing inappropriate answers. Known missing data structure data may arise due to the structure of the questionnaire. For example, the first question on a survey might be: 'If YES, skip to question 4', resulting in questions 2 and 3 missing. If the structure of the questionnaire is known, this type of missingness can be evaluated easily. However, if this information is not available, the mechanism responsible for producing missing data must be inferred from the data.

Another common source of known and unknown structured missingness is medical examination data. The results of particular medical tests may be: missing for purely random reasons, missing due to the procedure, or missing based on decisions arising from the observed data. For example, a young patient is young may not be subjected to neurodegenerative tests reserved for older workers. A final example is dropouts in a longitudinal study, where participants do not return for future testing sessions. In this case, it is difficult, sometimes impossible, to ascertain the reason for the dropouts, and hence, whether the missingness structure. However, this ascertainment is essential if the estimates based on these data are to be believed as unbiased, [@Simon1986; @Little1988; @Rubin1976].

Other categories of missing data are sometimes identified: Missing Completeley at Random (MCAR), Missing At Random (MAR), and Missing Not At Random (MNAR) [ @Little2002]. MCAR describes where missingness has no association with the observed or unobserved data. MAR describes cases where missingness depends on data observed, but not data unobserved. MNAR is where the missingness of the response is related to an unobserved value relevant to the assessment of interest.

XXX How have you met missings before? What examples of data were you working with that had missing value problems?

## Existing packages for handling missing data

XXX Describe exisiting packages, and group them into what they do

Software focussing on missing data typically focus on imputation or visualisation. Imputation packages such as mice, Hmisc, mi, Amelia, and mitools provide functions to facilitate imputation, aggregation of imputations, diagnostics, and define user-written imputation functions [@mice; @Hmisc; etc. ]. It is important to remember that care must be taken to avoid bias when performing multiple imputation.

Missing data visualisation packages include the R package VIM, and the stand alone softwares MANET, ggobi, and MissingDataGUI. 


MANET (Missings Are Now Equally Treated), provides univariate visualisations of missing data using linked brushing, using a referential plot to display the proportion of missingness for each variable as a filled bar plot, and then displayed plots of the data, using a histogram (for continuous data) or  a barplot (for categorical data). The plots of the data were highlighted when values of one variable were in the same row as a missing value. 

*some sort of figure describing this?*

ggobi provided multivariate visualisation of missing data using two methods. The first involved a reference plot providing a visual cross-tabulation of the data according to missing or not missing. This was then linked to a parallel coordinate plot, such that whichever values from the reference plot were then highlighted on the parallel coordinate plot. ggobi also provided scatterplot plots of missingness, where the variables are ...

XXX One sentence on why MissingDataGUI is too heavy weight for general purpose

MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually, it also provides visualisation methods for visualising imputated values. A limitation of MissingDataGUI is that it breaks workflow by being a separate part of analysis, where one might be performing an analysis using R, using Also being separate from 

XXX One paragraph on how ggplot2 handles missing values (or not): a message is printed and then the missings are ignored

ggplot2, a (very popular) package for producing elegant graphics in R, currently only provides visualisation for missing data when there is a category variable, this provides visualisations for missing data when the values being plotted are categories, it treats one of the categories as a NA value. Outside of this, ggplot2 does not provide much support for visualisation of missing data. It currently prints out a warning message when a plot us shown that contains missing data, which states that it has removed a particular number of rows containing "non-finite" values. 

See some examples below (not for the paper, but for explaining right now:

```{r ggplot-missing-vals}

data_test <- data_frame(x = c(1,2,3,4,5,5,5,5,NA,NA),
                        y = c(9,5,6,7,4,9,2,4,4,5))

  ggplot(data_test,
         aes(x = factor(x),
             y = y)) + 
  geom_bar(stat = "identity")

  ggplot(data_test,
         aes(x = factor(x),
             y = y)) + 
  geom_boxplot()

  ggplot(data_test,
         aes(x = factor(x),
             y = y)) + 
  geom_point()
```

# Data structures for missing data

```{r create-data-structures}

# create a simple example of missing data for the paper to illustrate the shadow matrix
df_example <-  tibble::tribble(~V1, ~V2, ~V3, ~V4,
                               "A", 15, 1.2, NA,
                               "A", NA, NA, T,
                               "A", 18, NA, F,
                               "B", 5, 1.6, T,
                               "B", NA, 0.7, T,
                               "B", 12, NA, F)

```


Representing missing data structure is achieed using the shadow matrix, introduced in Swayne and Buja [-@Swayne1998]. The shadow matrix is the same dimension as the data, and consists of binary indicators of "missingness" of data values. In our case, missing is represented as "NA", and not missing is represented as "!NA", although these may be represented as 1 and 0, respectively. This helps us explicitly describe the missingness structure. Representing it as its own separate matrix also helps us separate out the multivariate nature of missing data, as different cases have missing values in different sets of variables.

![diagram of shadow missing](/Users/tierneyn/Google Drive/ALL THE THINGS/PhD/code/R/jsm2017/diagram.pdf)

The shadow matrix is represented below:

```{r as-shadow,echo = FALSE}

knitr::kable(as_shadow(df_example))

```

It can also be bound in a wide format, column-wise, to the existing datastructure, adding the suffix "_NA" to the data.

```{r bind_shadows}

knitr::kable(bind_shadow(df_example))

```

Another format is to display it in long form, as below:

```{r gather_shadow}

as_shadow(df_example) %>%
  mutate(rows = 1:nrow(.)) %>%
  gather(key = "var",
         value = "miss",
         -rows) %>%
  slice(1:10) %>%
  knitr::kable()

```


## Visualisation of missing data

**Heatmap (plot of the long form)**

One common method for visualising missing data is to display a heatmap of the shadow matrix. This approach can be very helpful for giving an overview of which variables contain the most missingness. Methods can also be applied to rearrange rows and columns to find clusters, and identify other interesting features of the data that may have previously been hidden or unclear. This method is shown below using the `vis_miss` from the `visdat` package.

```{r }

library(visdat)
vis_miss(airquality)

```

Similar approaches has been used in other missing data packages such as VIM, mi, Amelia, and MissingDataGUI. However this plot provides the additional benefits of being in the ggplot framework, which gives users greater control over the plot appearance, the interface and provides the plot in ggplot, which allows for the user to alter and update the figure as they need. To produce this plot, the data is required to be in long form.

**Facetted plots**

An advantage of the wide shadow format is that it allows for referring to missingness of other variables along the values of another variable. For example:

```{r}

ggplot(data = bind_shadow(airquality),
       aes(x = Ozone)) + 
  geom_histogram() + 
  facet_wrap(~Solar.R_NA,
             ncol = 1)

ggplot(data = bind_shadow(airquality),
       aes(x = Ozone,
           colour = Solar.R_NA)) + 
  geom_density()

```

This plot shows a histogram of Ozone, showing what the distribution of Ozone is when Solar.R is missing.

Using this data structure facilitates missing data visualisation with ggplot as it means that the user can directly refer to the variable that they want to explore missingness. In the case above, the user is looking at a histogram of Ozone, but is then able to look at how many Ozone values are affected by Solar.R. In cases where there is no missing data in the variable that they want to "split" the missingness by, the plot simple returns a single facetted plot.

A similar method of visualisation could also be explored using `geom_missing_point()` from the ggmissing package:

```{r}

ggplot(data = airquality,
       aes(x = Ozone,
           y = Solar.R)) + 
  geom_missing_point()
```

This utilises methods from ggobi and Manet, where missing values are replaced with values 10% lower than the minimum value in that variable, the missing values are also different colours, so that missingness becomes preattentive.

## Usage

Numerical summaries of missing data are also made easy with some helper functions from the `ggmissing` package. For example, finding the overall proportion of missing values in the data overall, or the cases, or variables, can be done with `percent_missing_*` functions.

Single number summaries:

- The proportion elements in dataset that contains missing values
- The proportion of variables that contain any missing values
- the proportion of cases that contain any missing values


```{r}

percent_missing_df(airquality)
percent_missing_var(airquality)
percent_missing_case(airquality)

```

We can also look at the number and percent of missings in each case, and in each variable with `summary_missing_case`, and `summary_missing_var`, which both return tibbles.

```{r}

ggmissing::summary_missing_case(airquality)
ggmissing::summary_missing_var(airquality)

```

We can also present tabulations that present more complicated inferences with `table_missing_case` and `table_missing_var`. These tally up the number of missings in each case or variable, then describe how many cases or variables have that many missings, and the percentage of 

```{r}

ggmissing::table_missing_case(airquality)
ggmissing::table_missing_var(airquality)

```

<!-- it would be really cool if we could implement dplyr `group_by` syntax for the data, to produce summaries of missingness for 1993 and 1997 respectively. -->

# Discussion

Here we have discussed some of the common issues with investigating missing data.


You can also write a few sentences on computing on the fly vs data storage

- this is the `bind_shadow` versus computing on the fly with `is_na`

```

ggplot(data = airquality,
       aes(x = Ozone)) + 
  geom_histogram() + 
  facet_wrap(is_na(Solar.R),
             ncol = 1)

ggplot(data = airquality,
       aes(x = Ozone,
           colour = is_na(Solar.R)) + 
  geom_density()
```

It is important to consider then the implications for storage, when using bind_shadow, and computation, when using the `is_na` approach.

`bind_shadow` provides some flexibility and extensibility, as you could then add complexity to the kinds / flavours / mechanisms of missingness, extending from `NA`, `!NA`, and `NA_missing_for_reason_A`, `NA_missing_for_reason_B`, `NA_missing_for_reason_C`.


**Limitations**

- We didn't discuss how to visualise imputations, but you can imagine a similar framework for labelling imputations in a data structure.
- 

**Future work**

explore model-based approaches for exploring  based missing data structures.

- How does this expand the field?
- How is this different to previous findings / usage?

# to do

- Check write up of imputation / visualisation
- check vis writeup
- explain the figure of the shadow matrix
- Check the "usage" section.


# References