---
title: "Extra notes from JSM"
author: "Nicholas Tierney"
date: "15/12/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Appendix 

## ggplot's current approach to missing data


```{r ggplot-missing-vals}

data_test <- data_frame(x = c(1,2,3,4,5,5,5,5,NA,NA),
                        y = c(9,5,6,7,4,9,2,4,4,5))

data_test

  ggplot(data_test,
         aes(x = factor(x),
             y = y)) + 
  geom_bar(stat = "identity")

  ggplot(data_test,
         aes(x = factor(x),
             y = y)) + 
  geom_boxplot()

```


## Shadow Matrix Examples


```{r create-data-structures}

# create a simple example of missing data for the paper to illustrate the shadow matrix
df_example <-  tibble::tribble(~V1, ~V2, ~V3, ~V4,
                               "A", 15, 1.2, NA,
                               "A", NA, NA, T,
                               "A", 18, NA, F,
                               "B", 5, 1.6, T,
                               "B", NA, 0.7, T,
                               "B", 12, NA, F)

```

Example of the shadow matrix

```{r as-shadow,echo = FALSE}

knitr::kable(as_shadow(df_example))

```

Example of the augmented shadow matrix:

```{r bind_shadows}

knitr::kable(bind_shadow(df_example))

```

Example of the gathered shadow matrix:

```{r gather_shadow}

as_shadow(df_example) %>%
  mutate(rows = 1:nrow(.)) %>%
  gather(key = "var",
         value = "miss",
         -rows) %>%
  slice(1:10) %>%
  knitr::kable()

```


Other categories of missing data are sometimes identified: Missing Completeley at Random (MCAR), Missing At Random (MAR), and Missing Not At Random (MNAR) [@Little1988]. MCAR describes where missingness has no association with the observed or unobserved data. MAR describes cases where missingness depends on data observed, but not data unobserved. MNAR is where the missingness of the response is related to an unobserved value relevant to the assessment of interest. These categories will not be discussed further in this paper, and we will instead focus on the approaches used to handle missing data structure.

XXX How have you met missings before? What examples of data were you working with that had missing value problems?



# Extras

### Defining missing data and its various forms.

It is helpful to first define what missing data is, and what it is not. Missing data is data that we know should exist, but for some particular reason is not recorded. For example, if there is temperature data recorded every hour of every day, and one particular hour is missing on a particular day, this is missing data. This is contrasted to data that does not exist at all, for example, combining person-level data with environment level data - A person would not have an ambient temperature, and an environment does not have a pulse. These data are sometimes referred to as NULL data, or non-data. 

One of the motivations for understanding structure of missing data is to understand *why* it is missing in the first place. In the example above for the temperature data, the temperature might not have been recorded due to instrument failure, or system-wide shut down, perhaps it was scheduled for maintenance. 

This means that when answering the question "is this data entry filled?" one must actually consider three possible answers: "Yes", "No, but it can be", and "No, and it cannot be".

### Previous examples of missing data
For example, assessments of lung function taken at a workplace may be missing for workers who are on vacation. If there is no known or measurable relationship between the timing of the tests and the timing of vacations, and if the other relevant features of the workers who are on vacation at the time of the tests are similar to that of other workers, then these missing data can be considered MCAR. 

 For example if the missing lung function data occurs in workers who are being assessed for depression, and if there is no relationship between lung function and depression, then it can be considered as MAR.  For example, if BMI is of interest, but those with especially large BMIs are more likely to have missing BMI data, these data can be considered as MNAR. It is important for researchers to recognise MNAR as it introduces bias into the estimation of associations and parameters of interest. For example, if lung function and BMI are negatively correlated, an estimate of BMI based on the MNAR may be too low.

These three varieties of missing data could be further divided into a knowable structure (MAR) or an unknown structure (MAR or MNAR), where the process driving data becoming missing are either known or unknown, and structure refers to variables and interactions that may influence missingness. Data MCAR are without a missingness structure, as missingness does not have any dependence on other variables. Determining whether this is known or unknown is important for determining whether bias may be introduced into the analysis.

### on tests for missing data

Tests confirming whether data is MCAR or not are very useful as they open up the doors for the use of standard multiple imputation techniques.  As described by Little,who proposed a single test statistic for testing MCAR, involving an evaluation of equality of means between identified missing data groups. Rejection of this test result gives strong evidence that the data are not MCAR. Little's test of MCAR is widely used today, especially in social science8 and medical research.9 Recent research has also provided statistical tests and software that evaluate missing data via patterns, equality of means, and homogeneity of variance, and allow for non-normal data. 

This is achieved, for example, in the MissMech package for the R statistical software,10 which uses imputation (from either normal or non-normal distributions) to compare means and covariances. These tests enable the researcher to determine whether or not there is sufficient evidence for data to be declared as MCAR.  

However, understanding how and why missingness is being generated can become arduous when handling larger data sets, as they can have many missingness patterns, making inference difficult, as there are many combinations of missingness to explore. Reliance on statistical significance testing to assess whether data are missing may fail to address settings where there may not be significant missingness, but a complete case analysis may still result in bias (11). Approaches that elucidate missingness structure that are simple to understand and implement, are therefore still in demand.

### on other methods for missing data

Common methods of handling missing data, such as complete case analysis, missing indicator method, and last case carried forward have been shown to be acceptable when data is MCAR.12 ,13 That being said, most recommendations now are to use multiple imputation, but subject to some care as it only reduces bias from analysis when data are MAR or MCAR; multiple imputation also requires variables that influence missingness to be included in the imputation model.1–4 ,14 When data are MNAR, multiple imputation can be used but requires the MNAR mechanism to be known, which is not often undertaken in practice.3 Improving the understanding of missingness structure in a data set allows for consideration of other appropriate multiple imputation methods, or other methods to incorporate partially observed variables, such as random effect models, Bayesian methods, down-weighting analyses, or pattern mixture models.2 ,15 ,16

### missing data as a dimension

It creates a tension as there is this extra dimension to the data. 
Visualising missing data can be very challenging, and on the surface is somewhat paradoxical: how do you visualise things that are missing? 